A comparison of different methods for calculating tangent-stiffness matrices in a massively parallel computational peridynamics code.

### 2013-09-02 jacobian_calc_CMAME.tex
Cleaned up methods, results table

\subsubsection{Justification: Goals, Assumptions, and Metrics}

A goal of the study was to rank complex-step against forward-difference and central difference on
the basis of accuracy. AD was ommited from the comparison because it served as the standard of
accuracy for the other methods in the absence of appropriate analaytical forms for the Jacobian
associated with the nonlinear system used solved in the study. The assumption that AD is accurate
enough to serve as a standard is supported by AD's implementation as a computerized chain rule as
explained in \ref{ADsubsection}. For a concrete comparison of each of the methods including AD to a
known analytical Jacobian, refer to the corresponding author's website, specifically the C++
source-code example \emph{BeamAndSpringSystem}.

Because it makes no sense to compare tangent-stiffness matrices from different load steps, or from
different iterations, it was necessary to solve one load step and conclude each iteration within
that load step by updating guesses only from the the results of the AD method, and to subsequently
feed all four methods the same updated guess the following iteration. This decision precluded a
comparison of convergence rate, since if the methods were allowed to solve a problem at their
invidual pace, differences in accuracy would produce differences in guess updates and therefor the
number and nature of iterations perfomed. Different guesses from iterations started with different
previous guesses could not be data for a valid comparison of tangent-stiffness accuracy between
methods.  However a separate convergence rate comparison study could be done given some
modifications to the materials for the study described in this paper, and of course, the resources
to perform that study.  Instead, in this study, speed of iteration was measured because it could be
so done at the same time as Accuracy was being measured given a single tangent-stiffness
calculation, which was attractive due to limited resources.  Additionally, having each of the
methods physically operate in the same process, single or parallel, allowed the comparison of
tangent-stiffness matrices as loaded from random access memory rather than from the hard-disk,
giving the benefit of vastly greater speed and the benefit of simplicty compared to some other
solution speculatively involving dynamic file management on files which for one of the components of
the study would be on the order of $1XE11$ bytes, for which "vastly" is not hyperbole.

The other goal was to rank the four methods, that is including AD, on the basis of speed. It was
assumed that the order that each method was evaluated was unimportant, that evaluating each method
sucessively within each solver iteration did not affect their performance individually and that
speed of computation did not change with time. These assumptions allowed the test program to run the
same problem with each of the methods at effectively the same time and generate an equal volume of
data from each method. It was also assumed that the Jacobian matrix calculation time for each of the
methods did not vary based on what values the independent variables held, as they do from iteration
to iteration as the guessed equilibrium solution is updated. This assumption allowed calculation
time measurements to be averaged over all iterations within a solution attempt. The purpose of
averaging calculation time measurements was to informally address the extraneous variables of
evaluation order and computer system load due to other user's prcocesses or processes not associated
with the study.

The goals of collecting accuracy and speed measurements were achieved by developing and implementing
metrics within the simulation program used in the study. The metric used to measure the accuracy of
a tangent-stiffness matrix was the Frobenius norm of the element-wise difference between the
tangent-stiffness matrix produced by the method being evaluated and the tangent-stiffness matrix
produced by the AD based method, given that both methods were set upon the same problem. The lower
the value of this metric, the closer the other method's Jacobian was to the AD Jacobian, and
therefore the more accurate the method was. The expression for the accuracy metric:\\ $D =
\sqrt{\sum_{i=0}^{n-1}\sum_{j=0}^{n-1} (J_{AD}(i,j) - J_{M}(i,j))^2}$\\ Where $D$ is distance, n is
the number of degrees of freedom, $J_{AD}$ is the Jacobian matrix produced by the AD based method, M
stands for other \emph{M}ethod, and the root of the summed squared element-wise differences between
the matrices is taken.  The metric used to compare the speed of the methods was the time in seconds
taken to resume execution in the calling method after instructing the implementation of the method
being evaluated to run. Another basis of comparison, but not quite a metric, called computational
efficiency and used in the study is based on the ratio of calculation time to number of Jacobian matrix
elements.

\subsubsection{Materials: Computers, Other Software}

@@

\section{Results and Discussion}
\subsection{Speed Data}
\subsection{Accuracy Data}
\subsection{Efficiency Data}

@@

\section{Unfinished table}

\begin{center}
    \begin{tabular}{  p{1cm}  p{1cm}  p{1cm}  p{1cm}  p{2cm}  p{3cm} p{2cm} }
    \hline
    \multicolumn{7}{c}{Full Test Matrix} \\ \hline
    Cores & Target & Rows & horizon (m) & Nonzero Rows & Calc. Time (s) & Accuracy Diff. (MPa)\\ 
    \multicolumn{5}{c}{} & CS, CD, FD, AD & CS, CD, FD \\ \hline 
	1 & 1E3 & 3000 & .3 & 1994454  & 3.5, 2.7, 1.7, 1.6 & 1.92E-10, 1.21E-4, .137 \\ \hline
	1 & 2E3 & 5400 & .24 & 4252230  & 6.2, 4.9, 3.2, 2.9 & 2.28E-10, 9.94E-4, .148 \\ \hline
	1 & 3E3 & 8526 & .21 & 7792128  & 11.2, 8.9, 5.7, 5.2 & 2.38E-10, 1.59E-4, .145\\ \hline
	1 & 4E3 & 12288 & .19 & 15389352  & 26.7, 21.0, 13.4, 12.2 & 2.33E-10, 4.61E-4, .12 \\ \hline
	1 & 6E3 & 18468 & .16 & 20027466  & 28.1, 22.2, 14.4, 13.1 & 2.76E-10, 1.05E-3, .145 \\ \hline
	1 & 8E3 & 24000 & .15 & 31388724  & 47.6, 37.7, 24.2, 21.9 & 2.64E-10, 1.65E-3, .133 \\ \hline
	1 & 1.2E4 & 33396 & .131 & 40060026  & 55.6, 44.1, 28.4, 25.9 & 3.03E-10, 1.92E-3, .148 \\ \hline
	1 & 1.6E4 & 50700 & .12 & 83778480  & 138.9, 109.6, 70.2, 64.0 & 3.63E-10, 1.64E-3, .123 \\ \hline
	1 & 3.2E4 & 96768 & .095 & 165140640  & 277.3, 218.1, 139.4, 126.5 & 3.26E-10, 2.18E-3, .128 \\ \hline
	32 & 1E6 & 3E6 & .03 & 1666968710  & 336.1, 277.7, 200.6, 233.1 & 6.21E-10, 1.52E-2, .176 \\ \hline
	64 & 1E6 & 3E6 & .03 & 1666962752  & 169.9, 140.7, 102.0, 119.7 & 6.20E-10, 1.50E-2, .177 \\ \hline
	96 & 1E6 & 3E6 & .03 & 1666760396  & 114.7, 95.0, 69.1, 79.7 & 6.18E-10, 1.50E-2, .177 \\ \hline
	128 & 1E6 & 3E6 & .03 & 1666892660  & 86.4, 71.8, 52.4, 58.8 & 6.16E-10, 1.47E-2, .177 \\ \hline
    \end{tabular}
\end{center}

>>>>

\subsubsection{Justification: Goals, Assumptions, and Metrics} 
A goal of the study was to rank complex-step against forward-difference and central difference on
the basis of accuracy. AD was ommited from the comparison because it served as the standard of
accuracy for the other methods in the absence of appropriate analaytical forms for the Jacobian
associated with the nonlinear system solved in the study. The assumption that AD is accurate
enough to serve as a standard is supported by AD's implementation as a computerized chain rule as
explained in \ref{ADsubsection}. For a concrete comparison of each of the methods including AD to a
known analytical Jacobian, refer to the corresponding author's website, specifically the C++
source-code example \emph{BeamAndSpringSystem}.

Because it makes no sense to compare tangent-stiffness matrices from different problems, load steps,
or from different iterations, it was necessary to solve one load step and conclude each iteration
within that load step by updating displacement guess only from the the results of the AD method, and
to subsequently feed all four methods the same updated guess the following iteration. This decision
precluded a comparison of convergence rate, since if the methods were allowed to solve a problem at
their individual pace, differences in accuracy would produce differences in guess updates and
therefor the number and nature of iterations perfomed. Different guesses from iterations started
with different previous guesses could not be data for a valid comparison of tangent-stiffness
accuracy between methods.  However a separate convergence rate comparison study could be done given
some modifications to the materials for the study described in this paper, and of course, the
resources to perform that study.   Additionally, having each of the methods physically operate in
the same process, single or parallel, allowed the comparison of tangent-stiffness matrices as loaded
from random access memory rather than from the hard-disk.  Loading from RAM gives the benefit of
vastly greater speed and the simpler programming compared to some other solution speculatively
involving dynamic file management on files which for one of the components of the study would be on
the order of $1XE10$ bytes for which "vastly" is not hyperbole. However, the price of running the
methods together in the same process and avoiding the use of the hard-disk was that at least two
tangent-stiffness matrices had to be stored in RAM during simulation, which meant that special
high-memory compute nodes were needed for the $1E6$ peridynamic element problem series.

The other goal was to rank the four methods, that is including AD, on the basis of speed. Instead of
convergence rate being measured, speed of iteration was measured because it could be so done at the
same time as accuracy was being measured given a single tangent-stiffness calculation.  It was
assumed that the order that each method was evaluated was unimportant, that evaluating each method
sucessively within each solver iteration did not affect their performance individually and that
speed of computation did not change with time. These assumptions allowed the test program to run the
same problem with each of the methods at effectively the same time and generate an equal volume of
data from each method. It was also assumed that the Jacobian matrix calculation time for each of the
methods did not vary based on what values the independent variables held, as they do from iteration
to iteration as the guessed equilibrium solution is updated.  This assumption allowed calculation
time measurements to be averaged over all iterations within a solution attempt. The purpose of
averaging calculation time measurements was to informally address the extraneous variables of
evaluation order and computer system load due to other user's processes or processes not associated
with the study. It begs the question that if it is assumed that the value of independent variables
had no effect on iteration speed, than why couldn't each of the method have been run independently
so that convergence data were recorded, so long as system being solved was the same? The answer is
that this would have made accuracy comparisons invalid if measured concurrently with convergence
rate, while it would cost additional time and resources if convergence rate were measured in
serparate runs. Instead, while not part of the study, the example programs available on the
corresponding author's website allow the reader to make a comparison of the methods for themselves
on the basis of accuracy and convergence rate for two example nonlinear systems. In the examples,
notice how convergence rate decreases when the step size selected for forward-difference or
complex-step is set to larger numbers such that accuracy relative to AD is decreased. While likely
not the best possible implementations of these methods, they each were coded with the same skill
level such that a comparison of their relative strengths is reasonable from the standpoint of
intuition.

Once the selection of what to measure, why to measured it, and if it could be measured was
determined, the task was to find a way to take measurements that satisfied our goals. The goals of collecting accuracy and
speed measurements were achieved by developing and implementing metrics within the simulation
program used in the study. The metric used to measure the accuracy of a tangent-stiffness matrix was
the Frobenius norm of the element-wise difference between the tangent-stiffness matrix produced by
the method being evaluated and the tangent-stiffness matrix produced by the AD based method, given
that both methods were set upon the same problem. The lower the value of this metric, the closer the
other method's Jacobian was to the AD Jacobian, and therefore the more accurate the method was. The
expression for the accuracy metric:\\ $D = \sqrt{\sum_{i=0}^{n-1}\sum_{j=0}^{n-1} (J_{AD}(i,j) -
J_{M}(i,j))^2}$\\ Where $D$ is distance, n is the number of degrees of freedom, $J_{AD}$ is the
Jacobian matrix produced by the AD based method, M stands for other \emph{M}ethod, and the root of
the summed squared element-wise differences between the matrices is taken.  The metric used to
compare the speed of the methods was the time in seconds taken to resume execution in the calling
method after instructing the implementation of the method being evaluated to run and complete itself. Another basis of
comparison used in the study, but not quite a metric, called computational efficiency is based
on the ratio of calculation time to number of Jacobian matrix elements.

\subsubsection{Materials: Computers, Other Software}
Mesh generation for the test problems run in the study was done using the software package $Cubit$
developed at REFERENCE.

@@

\section{Results and Discussion}

\begin{center}
    \begin{tabular}{c c c c }
    \hline
    \multicolumn{4}{c}{Averaged Results, Entire Test Matrix} \\ \hline
    Cores & Jacobian Els. & Calc. Time (s) & Accuracy Diff. (MPa)\\ 
    \multicolumn{2}{c}{} & CS, CD, FD, AD & CS, CD, FD \\ \hline 
	1 & 1.99E6  & 3.5, 2.7, 1.7, 1.6 & 1.92E-10, 1.21E-4, .137 \\ \hline
	1 & 4.25E6  & 6.2, 4.9, 3.2, 2.9 & 2.28E-10, 9.94E-4, .148 \\ \hline
	1 & 7.79E6  & 11.2, 8.9, 5.7, 5.2 & 2.38E-10, 1.59E-4, .145\\ \hline
	1 & 1.54E7  & 26.7, 21.0, 13.4, 12.2 & 2.33E-10, 4.61E-4, .12 \\ \hline
	1 & 2.00E7  & 28.1, 22.2, 14.4, 13.1 & 2.76E-10, 1.05E-3, .145 \\ \hline
	1 & 3.14E7  & 47.6, 37.7, 24.2, 21.9 & 2.64E-10, 1.65E-3, .133 \\ \hline
	1 & 4.01E7  & 55.6, 44.1, 28.4, 25.9 & 3.03E-10, 1.92E-3, .148 \\ \hline
	1 & 8.39E7  & 138.9, 109.6, 70.2, 64.0 & 3.63E-10, 1.64E-3, .123 \\ \hline
	1 & 1.65E8  & 277.3, 218.1, 139.4, 126.5 & 3.26E-10, 2.18E-3, .128 \\ \hline
	32 & 1.67E10  & 336.1, 277.7, 200.6, 233.1 & 6.21E-10, 1.52E-2, .176 \\ \hline
	64 & 1.67E10  & 169.9, 140.7, 102.0, 119.7 & 6.20E-10, 1.50E-2, .177 \\ \hline
	96 & 1.67E10  & 114.7, 95.0, 69.1, 79.7 & 6.18E-10, 1.50E-2, .177 \\ \hline
	128 & 1.67E10  & 86.4, 71.8, 52.4, 58.8 & 6.16E-10, 1.47E-2, .177 \\ \hline
    \end{tabular}
\end{center}

\subsection{Speed Data}
\subsection{Accuracy Data}
\subsection{Efficiency Data}

@@
